Fitting the Undetected: Light-curve Feature Estimation in the Presence of Non-detections

### 2012-04-10 tex/ms.tex
revised manuscript, implementing changes discussed on 04/09 with hogg and jpl

@@

\section{Method}
\label{sec:method}

To recap, all we get at each epoch $t_i$ is a bit $q_i$ and, when
$q_i=1$, two data $f_i$ and $s_i^2$.  We don't get any information
about the upper limits or detection thresholds at the times $t_i$ when
$q_i=0$, and we don't know anything about the data quality (catalog
veracity) at any epoch.  For notational convenience we assemble all
the data into an enormous set $D$ given by
\begin{eqnarray}\displaystyle
D &\equiv& \{D_i\}
\\
D_i &\equiv& (q_i, f_i, s_i^2)
\quad ,
\end{eqnarray}
where the $f_i$ and $s_i^2$ data will be randomly generated or
meaningless at times at which the $q_i=0$.

We don't have access to all the information we need to use responsibly
the detections and non-detections in model fitting and inference.  We
have to instantiate latent variables to stand in for the missing
information.  For example, since we don't know the censoring threshold
or completeness limit for any observation $i$, we have to make it a
model parameter $b_i$.  Because we don't believe the reported
uncertainty variances $s_i^2$, we have to add parameters that model
the departure of the true uncertainty variance $\sigma_i^2$ from the
reported variance, and so on.

For the limited purposes here, the variable star will be periodic with
angular frequency $\omega$.  The expected brightness $\mu_i$ of the
star at time $t_i$ will be given by a linear combination of periodic
functions something like
\begin{eqnarray}\displaystyle
\mu_i &=& \sum_k A_k\,x_k(t_i|\omega)
\quad ,
\end{eqnarray}
where the $A_k$ are coefficients and the $x_k$ are functions.  This
can be generalized easily.  In addition to this expected brightness,
we assume that there is some either model uncertainty or stochastic
variation, leading to a \emph{model variance} $s_\mu^2$ at each point
(assumed constant but easily generalized).

If a source is detected at epoch $t_i$, we assume that its observed
flux $f_i$ is related to the expected flux $\mu_i$ by a Gaussian with
variance that has measurement and model contributions
\begin{eqnarray}\displaystyle
p(f_i|q_i,\sigma_i,\theta,I) &=& \left\{\begin{array}{ll}
  1 & \mbox{for $q_i=0$} \\
  N(f_i|\mu_i,s_\mu^2+\sigma_i^2) & \mbox{for $q_i=1$}
\end{array}\right.\label{eq:poff}
\\
\theta &\equiv& (\omega, \{A_k\}, s_\mu^2, \cdots)
\\
I &\equiv& (\{t_i\}, \mbox{assumptions})
\quad ,
\end{eqnarray}
where we have made the prior information about observation times $t_i$
and other assumptions explicit in the prior information blob $I$, and
we have a parameter blob $\theta$, which we will specify completely
below.  Note that the flux datum is ignored when the source is
undetected.

Similarly for the reported uncertainty variance $s_i^2$.  We don't
assume that it is correct in any case, but when the source is
detected, we assume that it is connected to the true uncertainty
$\sigma_i$ by a probabilistic likelihood, also Gaussian
\begin{eqnarray}\displaystyle
p(s_i|q_i,\sigma_i,\theta,I) &=& \left\{\begin{array}{ll}
  1 & \mbox{for $q_i=0$} \\
  N(s_i|\sigma_i,v_\sigma) & \mbox{for $q_i=1$}
\end{array}\right.
\\
\theta &\equiv& (\omega, \{A_k\}, s_\mu^2, v_\sigma, \cdots)
\quad ,
\end{eqnarray}
where we have added a model parameter $v_\sigma$, which represents the
variance in the distribution of reported uncertainties given the true
uncertainty; implicitly $v_\sigma$ is a component of $\theta$.  Again,
we are ignoring the variance datum when the source is undetected.

Under the assumption that the inclusion (or not) of the source in the
catalog is based on (or very strongly related to) the measured flux
(rather than the true flux), the natural likelihood for the bit $q_i$,
on which the above likelihoods are conditioned is
\begin{eqnarray}\displaystyle
P(q_i|b_i,\sigma_i,\theta,I) &=& \left\{\begin{array}{ll}
  \displaystyle\int_{-\infty}^{b_i} N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=0$} \\
  \displaystyle\int_{b_i}^\infty N(f|\mu_i,s_\mu^2+\sigma_i^2)\,\dd f & \mbox{for $q_i=1$}
\end{array}\right.\label{eq:pofq}
\quad,
\end{eqnarray}
where the integrals end at an observed-flux limit $b_i$, which (as
mentioned above) is also a model parameter.  The motivation for this
expression is that it imagines all possible observations $f$ of the
star at time $t_i$ given the model and uncertainties, and finds the
fraction that would make the (imagined) detection cut.

We are going to product together these likelihoods (for $q_i$, $f_i$,
and $s_i^2$); this looks like it is using the flux twice, once in the
PDF for $f_i$ in expression (\ref{eq:poff}) and once in the
probability for $q_i$ in expression (\ref{eq:pofq}).  It is not,
however, because in the probability for $q_i$ the flux $f$ that
appears is an integration variable, not an observation or datum.  We
treat the detections or non-detections as being ``prior'' to the
measurement.  This is the weakest plank in our otherwise impenetrable
fortress of solitude.

We are going to treat the true detection limit $b_i$ for each
observation, and its true uncertainty $\sigma_i$---two parameters
\emph{per data point}---as nuisance parameters.  We want to
marginalize them out.  This requires priors on these.  Again we choose
Gaussians for simplicity, but let the means and variances of these
Gaussians be new model parameters---hyperparameters if you wish.
\begin{eqnarray}\displaystyle
p(D_i|\theta,I) &=&
  \int P(D_i|b_i,\sigma_i,\theta,I)\,p(b_i|\theta)\,p(\sigma_i|\theta)\,\dd b_i\,\dd\sigma_i
\\
p(D_i|b_i,\sigma_i,\theta,I) &=&
  P(q_i|b_i,\sigma_i,\theta,I)\,p(f_i|q_i,\sigma_i,\theta,I)\,p(s_i|q_i,\sigma_i,\theta,I)
\\
p(b_i|\theta) &=& N(b_i|B,V_B)
\\
p(\sigma_i|\theta) &=& N(\sigma_i|S,V_S)
\\
\theta &\equiv& (\omega, \{A_k\}, s_\mu^2, v_\sigma, B, V_B, S, V_S)
\quad,
\end{eqnarray}
where we have (at last) explicitly assembled all the variable-star
parameters and hyperparameters into the big parameter vector $\theta$.
The huge sets of nuisance parameters $\{b_i\}$ and
$\{\sigma_i\}$ don't remain, because by integration we
removed them.

Finally and it goes without saying that if we treat the data points as
independent (as we are free to do, given the non-awesomeness of the
software data source), we have that
\begin{eqnarray}\displaystyle
p(D|\theta,I) &=& \prod_i p(D_i|\theta,I)
\quad.
\end{eqnarray}
This is the likelihood for variable-star parameters given the entire
data set (all the measurements and non-detections of this star from
all the epochs, as delivered by the untrustworthy robots).  It
``correctly'' or at least ``justifiably'' uses all of the information
available, without making strong assumptions about the survey or its
veracity.

@@

>>>>

\newcommand{\fobs}{f_i}
\newcommand{\sobs}{s^2_i}

\newcommand{\ftrue}{f_i^*}

@@

\section{Formulating the Model}
\label{sec:model}

%This note, based on the original note from JWR on Feb. 8, 2012, includes revisions from David Hogg, James Long, Nat Butler, and Josh Bloom.

As a reminder, we are modeling light curve data in the presence of non-detections, which are epochs of observation in which no detection of the source of interest was made.

\subsection{Preliminaries}
\label{ss:prelim}

For each astronomical object, a photometric survey measures a multi-epoch light curve over $N$ (typically unevenly spaced) epochs.  At each epoch $i$, with associated time $t_i$, the survey takes an exposure at the location of the object and either (a) detects the object and records an estimate of its photon flux, $\fobs$ and the  variance of the statistical uncertainty in that estimate, $\sobs$, or (b) fails to detect the object.  In the latter case, most modern surveys either record a reference value to signify that no detection was made (as is done in ASAS) or an estimate of the upper detection limit, $b_i$, which is the brightest the object could have been given that it was not detected at a significant level by the software.

There are many reasons that a source might not show up in a catalog.  These include:
\begin{itemize}
\item low S/N of the source, due either to a higher noise level or a fainter signal,
\item the source falling outside of the detection window (e.g., near chip edge),
\item occulting of the source by an artifact of the detector (e.g., hot pixels, masked out), or
\item the source was out-shone by an intervening object (e.g., asteroid, comet, variable star, airplane, etc.).
\end{itemize}
Here, we present a statistical model that can be used to detect variable sources and model their variability using multi-epoch light curves containing epochs of non-detection.  Previous efforts to detect and model variability using multi-epoch photometry have typically ignored non-detections (REFs) or used them in an ad-hoc manner lacking statistical rigor (REFs).  An exception is \citet{2009AJ....137.4400L}, who measure proper motions of sources in SDSS falling below the detection limit.

In this paper, we will assume that $\fobs$ and $\sobs$ take on real-valued numbers in epochs for which the source was detected and receive the reference value {\tt NA} in epochs where no detection was made. For notational convenience we assemble all the light curve data for one source into a data set $D$ given by
\begin{eqnarray}\displaystyle
D &\equiv& \{D_i\}_{i=1}^N
\\
D_i &\equiv& (\fobs, \sobs)
\quad ,
\end{eqnarray}
where $(\fobs, \sobs)$ are the light curve measurements  at $t_i$.  The goal, for each astronomical object, is to construct the likelihood for the data $D$, given a set of model parameters that describe the variability of the object and the characteristics of the observations, and then to maximize the likelihood with respect to the model parameters.% or to use it in further inference.

\subsection{The Light Curve Model}

To model the mean brightness of the light curve as a function of time, we use a multiple-harmonic Fourier model with angular oscillation frequency $\omega$,
\begin{eqnarray}\displaystyle
\mu_i &=& A_0 + \sum_{k=1}^K A_k\, \sin (t_i \, \omega  k) + B_k\, \cos (t_i \, \omega  k)
\quad ,\label{eq:fourier}
\end{eqnarray}
where $\mu_i$ is the flux at time $t_i$, $\sqrt{A_k^2 + B_k^2}$ is the amplitude of the $k^{\rm th}$ harmonic of the frequency $\omega$ and $\tan^{-1}(B_k,A_k)$ is the relative phase offset of harmonic $k$.  The number of harmonics, $K$, can either be fixed or treated as a free parameter over which to optimize the likelihood.    In addition to the expected brightness in Equation \ref{eq:fourier}, we assume that there is uncertainty or inappropriateness in the model, leading to a \emph{model variance} $s_\mu^2$ at each point (assumed constant but easily generalized).

Additionally, we need to instantiate a latent variable, $b_i$, to represent the detection limit, in units of flux, at epoch $i$.  The $b_i$ parameter is essential because it constrains the possible values of the mean brightness, $\mu_i$, of the light curve when there is a non-detection.  By employing a hierarchical model for the distribution of $b_i$, we can fully utilize all of the information encoded in both the detections and non-detections when computing the data likelihood.   At each epoch, we connect the observed flux, $\fobs$, to the latent variable, $\ftrue$, signifying the true observable flux of the object, via the detection limit, $b_i$ by
\begin{eqnarray}\displaystyle
\fobs &=& \left\{\begin{array}{ll}
  \ftrue & \mbox{if $\ftrue \ge b_i$} \\
  \texttt{NA} & \mbox{if $\ftrue < b_i$} \\
\end{array} \right.
\end{eqnarray}
so that the observed flux is \texttt{NA} only when the true observable flux, $\ftrue$, is below the detection limit, $b_i$.

Also, because we do not necessarily believe the reported measurement uncertainties, $\sobs$, we choose to introduce a parameter, $\sigma^2_i$, to represent the true uncertainty variance for the brightness measurement at epoch $i$.

Hence, our initial model consists of the parameter vectors $\{f^*_1,...,f^*_N\}$, $\{b_1,...,b_N\}$ and $\{\sigma^2_1,...,\sigma^2_N\}$, and the model parameters
\begin{eqnarray}\displaystyle
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, \cdots) \quad ,
\end{eqnarray}
along with prior information about observation times, $\{t_1,...,t_N\}$ and other prior assumptions \joey{we should be more explicit here}, which we make explicit by creating the prior information set
\begin{eqnarray}\displaystyle
I &\equiv& (\{t_i\}_{i=1}^N, \mbox{assumptions}) \quad .
\end{eqnarray}
Our goal is to write down the form of the likelihood of the data, $D$, given $\theta$ and $I$.  Then, for each light curve we can search for the vector $\theta$ that maximizes the data likelihood, and use those maximum likelihood estimates for downstream astrophysical inference.

\subsection{Statistical Model for Light Curves with Non-Detections}

We model the observed flux, $\fobs$, as a Gaussian distribution with variance that has both measurement ($\sigma^2_i$) and model ($s_{\mu}$) contributions.  In the case that a detection is made, we require that the observed brightness be greater than the brightness of the detection limit ($\fobs = \ftrue \ge b_i$) while in the case that no detection is made, we require that the brightness (if it could be measured) be less than the detection limit ($\ftrue < b_i$).  To derive the likelihood of the observed $\fobs$, given $\sigma^2_i$, $\theta$ and $I$, we must integrate over the prior distribution of the unknown $b_i$,
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I) &=& \left\{\begin{array}{ll}
  N(\ftrue | \mu_i,  \sigma^2_i + s_{\mu}^2)\,  \int_0^{\ftrue} p(b_i | \theta)\, \dd b_i & \mbox{if $\fobs \ne \texttt{NA}$} \\
  \int_{0}^{\infty} \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma^2_i + s_{\mu}^2)\, p(b_i | \theta)\, \dd \ftrue\, \dd b_i & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik}
\\
p(b_i|\theta) &=& N(b_i|B,V_B)
\label{eq:bprior}
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, B, V_B, \cdots) \quad ,
\end{eqnarray}
where we have introduced the hyperparameters $B$ and $V_B$ for the Gaussian prior distribution of $b_i$.  In Equation \ref{eq:mlik}, in the epochs for which a detection was made ($\fobs \ne \texttt{NA}$), we marginalize over the unknown $b_i$ from $0$ (low brightness limit) to $\ftrue$, enforcing that the detection limit be fainter than the observed brightness.  Likewise, in the epochs for which no detection was made ($\fobs = \texttt{NA}$), we integrate the joint $(\ftrue, b_i)$ likelihood over all possible values of $b_i$ and over the unknown $\ftrue$ from $0$ to $b_i$, ensuring that the brightness (if it were able to be observed) be fainter than the detection limit.

In the above, we have assumed no extra information on each of the $b_i$ values besides the knowledge of whether a detection was made on that epoch.  Hence, we draw, in Equation \ref{eq:bprior}, each $b_i$ value from a global prior distribution which is the same at all epochs.  If instead, we are given an estimate of $b_i$ plus its error distribution for each epoch (which, in principle can be inferred from the raw telescope images), we can replace Equation \ref{eq:bprior} with a different distribution per epoch.  In the case that the $b_i$ are assumed to be completely known (without error), the data likelihood of $\fobs$ becomes
\begin{eqnarray}\displaystyle
p(\fobs |\sigma^2_i,\theta,I, b_i) &=& \left\{\begin{array}{ll}
  N(\ftrue | \mu_i,  \sigma^2_i + s_{\mu}^2)\,  I(\ftrue \ge b_i) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 \int_{0}^{b_i} N(\ftrue | \mu_i, \sigma_i^2 + s_{\mu}^2)\, \dd \ftrue & \mbox{if $\fobs = \texttt{NA}$} \\
\end{array}\right.\label{eq:mlik_s}
\end{eqnarray}
where the boolean indictor function, $I(\ftrue \ge b_i)$, is 1 if $\ftrue \ge b_i$ and 0 otherwise.

Next, we model the reported variance, $\sobs$, on the uncertainty of the magnitude measurement.  Instead of assuming that $\sobs$ is a perfect, error-free measurement, we probabilistically connect it to the true uncertainty variance, $\sigma^2_i$ through a Gamma likelihood, which is defined to take on non-negative values.  Our likelihood of observed $\sobs$, given $\sigma^2_i$, $\theta$ and $I$, is
\begin{eqnarray}\displaystyle
p(\sobs | \fobs, \sigma^2_i, \theta,I) &=& \left\{\begin{array}{ll}
  \Gamma (\sobs | \sigma^2_i, V_{\sigma} ) & \mbox{if $\fobs \ne \texttt{NA}$} \\
 1& \mbox{if $\fobs = \texttt{NA}$}
\end{array}\right.\label{eq:slik}
\end{eqnarray}
where $\Gamma(x | m, V)$ is the standard Gamma distribution with mean $m$ and variance $V$, 
and we have added a model parameter $V_\sigma$ to represent the
variance in the distribution of reported uncertainties given the true
uncertainty.  (Typically, the Gamma distribution is parameterized by parameters $(\alpha, \beta)$, where the mean is $\alpha\beta$ and the variance is $\alpha\beta^2$.)  The  purpose of 
 including the likelihood in Equation \ref{eq:slik} to the model, in practice, to keep the $\sigma^2_i$ from
drifting very far away from the $s^2_i$, as set by the hyperparameter
$V_\sigma$.

Putting it all together, we can write down the likelihood of the data, $D_i$, for the parameters $\theta$, on a single epoch, as
\begin{eqnarray}\displaystyle
p(D_i|\theta,I) &=& \int_0^{\infty} p(\fobs |\sigma^2_i,\theta,I)\, p(\sobs | \fobs, \sigma^2_i,\theta,I)\, p(\sigma^2_i | \theta, I)\,\dd \sigma^2_i
\\
p(\sigma^2_i|\theta, I) &\propto& \sigma_i^{-1}
\\
\theta &\equiv& (\omega, A_0, \{A_k, B_k\}_{k=1}^K, s_\mu^2, B, V_B, V_{\sigma}) \quad ,
\end{eqnarray}
where we have inserted the expressions from Equations (\ref{eq:mlik}) and (\ref{eq:slik}), and integrated out the nuisance parameter, $\sigma^2_i$.  We have assumed a Jeffrey's prior on $\sigma^2_i$, which is non-informative and invariant to reparametrization of the variance.  We could alternatively use an inverse-gamma prior, which takes two hyperparameters (and is conjugate in the case of a normal likelihood with unknown variance). % We have also assumed that $\fobs$ and $\sobs$ are conditionally independent given $\sigma^2_i$, $\theta$, and $I$.  This is a reasonable assumption in the case that a detection is made at epoch $t_i$.

Finally if we assume that the data collected at each epoch are independent given the model parameters, we have that
\begin{eqnarray}\displaystyle
p(D|\theta,I) &=& \prod_i p(D_i|\theta,I)
\quad.
\end{eqnarray}
This is the likelihood for the entire
data set (all the measurements and non-detections of this star from
all the epochs, as delivered by the survey) given the $2K + 5$ parameter vector $\theta$.  This model
``correctly'' or at least ``justifiably'' uses all of the information
available, without making strong assumptions about the survey or its
veracity.

@@

\joey{Note: ASAS provides periods and amplitude for all of the objects in ACVS!  We can and should use this as a comparison set}

\joey{From a brief reading of a few Mira papers, it seems that some papers throw away Miras for which no trough of the LC was observed.  This seems silly.  I have an idea of the P-A plots from Hipparcos.  Should be a good comparison set.  Also, we can consider applying this to Hipparcos, which has publicly available data.}
